Isomorphisms as Groupoids: \\ Homotopy Type Theory and Fractional Types

### 2013-10-19 f.tex
discussion about treating functions as collections of paths if these functions are isos
braids etc

\newcommand{\ssep}{\hspace{1em}} 
\newcommand{\sym}{c}       % braid: could be \sigma or c or \gamma
\newcommand{\cp}{\circ}                 % composition
\newcommand{\symi}{\sym\inv}
\newcommand{\inv}{^{-1}}
\newcommand{\id}{\textrm{\rm id}}       % identity
\newcommand{\x}{\otimes}
\newcommand{\boolt}{\mathit{bool}}
\newcommand{\ffv}{\mathit{false}}
\newcommand{\ttv}{\mathit{true}}
\newcommand{\notb}{\mathit{not}}
\newcommand{\union}{\cup}
\newcommand{\transport}[2]{\overline{#1}(#2)}
\newcommand{\refl}{\mathit{refl}}
\renewcommand{\path}{\leadsto}
\newcommand{\paths}[1]{\mathds{P}(#1)}
%% \renewcommand{\circ}{\circlearrowleft}
\newcommand{\alt}{~|~}
\newcommand{\dt}[1]{\llbracket #1 \rrbracket}
\newcommand{\leftv}[1]{\textsf{left}~#1}
\newcommand{\rightv}[1]{\textsf{right}~#1}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\identlp}{\mathit{identl}_+}
\newcommand{\identrp}{\mathit{identr}_+}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{identl}_*}
\newcommand{\identrt}{\mathit{identr}_*}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\distz}{\mathit{dist}_0}
\newcommand{\factorz}{\mathit{factor}_0}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\idc}{\mathit{id}}
\newcommand{\wedgesum}{\vee}
\newcommand{\smashproduct}{\wedge}
\newcommand{\symc}[1]{\mathit{sym}~#1}
\newcommand{\proves}{\vdash}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2\\\end{array}}
{\begin{array}{l}#3\\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}

@@

%% the next 7 figures are taken from Selinger's paper on graphical languages
%% for monoidal categories; if we use them make sure it's ok with him and
%% acknowledge him

In homotopy type theory, there is a focus on isomomorphisms represented by
paths (proofs of identities) and yet arbitrary functions are considered. We
propose to only consider functions that are themselves isomorphisms: this
will allow us to view the function itself as a collection of paths and it
doesn't lose anything in terms of computability because we know that we could
do everything with reversible functions if we wanted to.

Consider the function $\notb$ from the space $\boolt$ to itself. We can
represent this function as a pair of paths. Since we are in a topological
world, we should care whether the path top pass crosses over the bottom one
or vice-versa. In the diagram below, we chose for the top wire to cross over
the bottom one:

\begin{center}
  \begin{tabular}{@{}llc@{}}
    $\notb : \boolt \rightarrow \boolt$ & & 
    $\vcenter{\wirechart{@C=1.5cm@R=0.8cm}{
        *{}\wireright{r}{\ffv}&\blank\wirecross{d}\wireright{r}{\ffv}&\\
        *{}\wireright{r}{\ttv}&\blank\wirebraid{u}{.3}\wireright{r}{\ttv}&
        }}$ 
  \end{tabular}
\end{center}

It is important also to note that the two paths must be taken together
because there is a constraint between them. If the top path takes $\ffv$ to
$\ttv$ then the bottom path cannot take $\ttv$ to $\ttv$: it \emph{has} to
lead to $\ffv$. We will return to this point later.

For now, consider the functions:
\[\begin{array}{rcl}
\mathit{id}^1 &=& \lambda x.x \\
\mathit{id}^2 &=& \lambda x. \notb~(\notb~x)
\end{array}\]
These two functions are evidently extensionally equivalent and hence in the
conventional treatment of homotopy type theory, it is acceptable to assume
they are also propositionally equivalent, i.e., it is acceptable to postulate
a path $\mathit{id}^1 \path \mathit{id}^2$ in the space $\boolt \rightarrow
\boolt$.

When represented as paths, we get the following situation for the second
function:
\[\vcenter{\wirechart{@C=1.5cm@R=0.8cm}{
    *{}\wireright{r}{\ffv}&
    \blank\wirecross{d}\wireright{r}{\ttv}&
    \blank\wirecross{d}\wireright{r}{\ffv}&
    \\
    *{}\wireright{r}{\ttv}&
    \blank\wirebraid{u}{.3}\wireright{r}{\ffv}&
    \blank\wirebraid{u}{.3}\wireright{r}{\ttv}&
    \\
    }}
\]
Repeating the $\notb$ operation results in a twisted path that externally
behaves like the identity but that evidently not identical to the straight
path with no twists. Thus although one can justify that the two functions
should have a path between them, the function extensionality axiom does not
give insights into \emph{what} this path is actually doing topologically.

@@

Talk about groupoid cardinality and give an example of how the cardinality of
a group of order $n$ is $\frac{1}{n}$.

>>>>

\newcommand{\br}{\mathit{Br}}
\newcommand{\ssep}{\hspace{1em}} 
\newcommand{\sym}{c}       % braid: could be \sigma or c or \gamma
\newcommand{\cp}{\circ}                 % composition
\newcommand{\symi}{\sym\inv}
\newcommand{\inv}{^{-1}}
\newcommand{\lolli}{\multimap} 
\newcommand{\id}{\textrm{\rm id}}       % identity
\newcommand{\x}{\otimes}
\newcommand{\boolt}{\mathit{bool}}
\newcommand{\ffv}{\mathit{false}}
\newcommand{\ttv}{\mathit{true}}
\newcommand{\notb}{\mathit{not}}
\newcommand{\union}{\cup}
\newcommand{\transport}[2]{\overline{#1}(#2)}
\newcommand{\refl}{\mathit{refl}}
\renewcommand{\path}{\leadsto}
\newcommand{\paths}[1]{\mathds{P}(#1)}
%% \renewcommand{\circ}{\circlearrowleft}
\newcommand{\alt}{~|~}
\newcommand{\dt}[1]{\llbracket #1 \rrbracket}
\newcommand{\leftv}[1]{\textsf{left}~#1}
\newcommand{\rightv}[1]{\textsf{right}~#1}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\identlp}{\mathit{identl}_+}
\newcommand{\identrp}{\mathit{identr}_+}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{identl}_*}
\newcommand{\identrt}{\mathit{identr}_*}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\distz}{\mathit{dist}_0}
\newcommand{\factorz}{\mathit{factor}_0}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\idc}{\mathit{id}}
\newcommand{\wedgesum}{\vee}
\newcommand{\smashproduct}{\wedge}
\newcommand{\symc}[1]{\mathit{sym}~#1}
\newcommand{\proves}{\vdash}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2\\\end{array}}
{\begin{array}{l}#3\\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}

@@

%% the next figures are taken from Selinger's paper on graphical languages
%% for monoidal categories; if we use them make sure it's ok with him and
%% acknowledge him

In homotopy type theory, there is a focus on isomorphisms: for most of the
spaces an isomorphism between $A$ and $B$ is represented by a path denoting a
particular proof of the identify $A \equiv B$. Different proofs yield
different paths and these proofs can be related by 2-paths to reason about
proof equivalence and so on. Yet in the middle of all these identities and
isomorphisms, conventional homotopy type theory deals with arbitrary
functions, then defines what it means \emph{extensionally} for functions to
be isomorphic, and \emph{then} postulates an axiom that extensionally
equivalent functions are identical in the sense of having a path between
them. This treatment gives no insight about the nature paths between
functions and is undesirable for many reasons.

We propose to just consider functions are isomorphisms by construction. This
doesn't lose anything in terms of computability because we know that we could
do everything with reversible functions if we wanted to. There are actually
many advantages of doing so as we and others have argued at length
before. Anyway, the point is that if we have isomorphisms-by-construction
functions then the function itself becomes naturally a collection of
paths. We illustrate this point with examples.

Consider the function $\notb$ from the space $\boolt$ to itself. We can
represent this function as a pair of paths. Since we are in a topological
world, we should care about how the paths cross over each other.  In the
diagram below, we chose for the top path to cross over the bottom one:
\begin{center}
  \begin{tabular}{@{}llc@{}}
    $\notb : \boolt \rightarrow \boolt$ & & 
    $\vcenter{\wirechart{@C=1.5cm@R=0.8cm}{
        *{}\wireright{r}{\ffv}&\blank\wirecross{d}\wireright{r}{\ffv}&\\
        *{}\wireright{r}{\ttv}&\blank\wirebraid{u}{.3}\wireright{r}{\ttv}&
        }}$ 
  \end{tabular}
\end{center}
It is important to note that the two paths must be taken together
because there is a constraint between them. If the top path takes $\ffv$ to
$\ttv$ then the bottom path cannot take $\ttv$ to $\ttv$: it \emph{has} to
lead to $\ffv$. In other words, it is incorrect to focus only on the top path
and conclude that there is an identity between $\ffv$ and $\ttv$. What is
correct is that, collectively, there is an identity between $\boolt$ and
$\boolt$ that permutes the elements of $\boolt$. Technically we have a
\emph{braid}.

We will return to this point later. But for now, consider the functions:
\[\begin{array}{rcl}
\idc^1 &=& \lambda x.x \\
\idc^2 &=& \lambda x. \notb~(\notb~x)
\end{array}\]
These two functions are evidently extensionally equivalent and hence in the
conventional treatment of homotopy type theory, they are also treated as
propositionally equivalent by the function extensionality axiom. When the
functions themselves are represented as paths, however, we get the following
situation for the second function:
\[
\idc^2 = \lambda x.\notb~(\notb~x) \qquad
\vcenter{\wirechart{@C=1.5cm@R=0.8cm}{
    *{}\wireright{r}{\ffv}&
    \blank\wirecross{d}\wireright{r}{\ffv}&
    \blank\wirecross{d}\wireright{r}{\ffv}&
    \\
    *{}\wireright{r}{\ttv}&
    \blank\wirebraid{u}{.3}\wireright{r}{\ttv}&
    \blank\wirebraid{u}{.3}\wireright{r}{\ttv}&
    \\
    }}
\]
Repeating the $\notb$ operation results in a twisted braid that extensionally
behaves like the trivial braid but that is evidently not identical to
it. Formally to find a (2-)path between the functions requires us to find an
\emph{isotopy} between the two braids, where loosely speaking an isotopy is
essentially homotopy in the case of bijections. The two braids for $\idc^1$
and $\idc^2$ are \emph{not} isotopic so there is no path between them in our
world. What is interesting is that the braid isotopy problem has efficient
solutions and hence we are in a situation where function equivalence has
efficient algorithms.

Note that one can define another version of boolean negation $\notb'$ in
which the bottom path is the one that crosses over the top one. Using this
version, the function $\idc^3 = \lambda x.\notb'~(\notb~x)$ would look like:
\[
\idc^3 = \lambda x.\notb'~(\notb~x) \qquad
\vcenter{\wirechart{@C=1.5cm@R=0.8cm}{
     *{}\wireright{r}{\ffv}&
     \blank\wirecross{d}\wireright{r}{\ffv}&
     \blank\wirebraid{d}{.3}\wireright{r}{\ffv}&
     \\
     *{}\wireright{r}{\ttv}&
     \blank\wirebraid{u}{.3}\wireright{r}{\ttv}&
     \blank\wirecross{u}\wireright{r}{\ttv}&
     \\
     }}
\]
And this version is isotopic to the trivial braid, and hence we would have a
(2-)path between $\idc^1$ and $\idc^3$.

To summarize, if we restrict our attention to a certain class of functions,
these functions are naturally described as braids and function equivalence
(i.e, paths between functions) are generated by braid isotopy. This is
already quite neat but note that we have only talked about functions
\emph{between} spaces and modeled these as braids. What we have not done yet
is describe the space of functions itself. In our running example, we need to
describe some space $\boolt \lolli \boolt$ that represents the various
functions we discussed from $\boolt$ to $\boolt$.

%%%%%%%%%%%%%%%%%
\subsection{Pointed Spaces}

We will construct this in steps. First we note that a function is a ``large''
object that can be applied to many possible arguments. Each application is
however associated with exactly one argument. Thus, in this first step, we do
not model functions in their entirety; we only model functions \emph{at one
  particular point}. We will return to the issue of modeling the entire
function object later. For now, we have technically moved from \emph{spaces}
to \emph{pointed spaces}. Let's revisit our previous discussion from the
perspective of pointed spaces.

The type $\boolt$ is now split in two types: the type $(\boolt,\ffv)$ with
$\ffv$ as the distinguished point and $(\boolt,\ttv)$ with $\ttv$ as the
distinguished point. We now have two copies of the function $\notb$, a
version $\notb_{\ffv} : (\boolt,\ffv) \rightarrow (\boolt,\ttv)$ and a
version $\notb_{\ttv} : (\boolt,\ttv) \rightarrow (\boolt,\ffv)$ depending on
the starting basepoint. Indeed functions must respect the basepoints, i.e.,
map basepoints to basepoints. The function is still a braid on the complete
$\boolt$ space but in the first case, the distinguished point $\ffv$ tells us
that the function is ``applied'' to $\ffv$ and hence that the top branch is
the one in focus.

Concretely then we want spaces (i.e., groupoids) $(\boolt,\ffv) \lolli
(\boolt,\ffv)$, $(\boolt,\ffv) \lolli (\boolt,\ttv)$, $(\boolt,\ttv) \lolli
(\boolt,\ffv)$, and $(\boolt,\ttv) \lolli (\boolt,\ttv)$. As examples, the
first space will contain $\idc^1_{\ffv}$ and $\idc^3{\ffv}$; the second space
will contain $\notb_{\ffv}$ and $\notb'_{\ffv}$; the third space will contain
$\notb_{\ttv}$, and the last space will contain $\idc^1_{\ttv}$. In the first
space, there will be a path between $\idc^1_{\ffv}$ and $\idc^3{\ffv}$ as
there is an isotopy between them etc. To describe these spaces fully, we need
introduction rules, elimination rules, computation rules, and induction
principles, etc. It should be clear that we can't just use
$\lambda$-expressions as introduction rules: they are not refined enough to
capture all the invariants we want.

Let's start with some basic observations. How many points (i.e., functions or
braids) are there in the space $(\boolt,\ffv) \lolli (\boolt,\ffv)$. These
are all the twists on two elements with the paths starting and ending at
$\ffv$ highlighted: by giving a twist in one direction the value $+1$ and a
twist in the other direction the value $-1$, we see that this is isomorphic
to the integers under addition. We would like to argue that this space has
groupoid cardinality 1 and if we succeed then it is natural to think of it as
the product of a space with fractional cardinality $\frac{1}{2}$ and the
space $\boolt$ with cardinality 2. Think of all the words over the alphabet
$\{ 0, 1, -1\}$ with rules $0x\rightarrow x$, $x0\rightarrow x$,
$1(-1)\rightarrow\epsilon$ and $(-1)1\rightarrow\epsilon$. This produces
equivalence classes which describe the connected components (isotopy
equivalences) of the space $\boolt\lolli\boolt$. Let's denote by $V_n$ the
size of the equivalence class of words whose value is $n$ where the value is
the sum of the elements. The groupoid cardinality of the space is therefore
$\Sigma_{n\in\mathbb{Z}} \frac{1}{A_n}$. Does this thing have value 1???
Plausible. Assume yes for now and proceed under the assumption that it is 1
and hence that it makes sense to decompose the space into the product of
something of cardinality $\frac{1}{2}$ and something of cardinality $2$. So
what kind of groupoid would have cardinality $\frac{1}{2}$. This actually is
something well studied. Talk about groupoid cardinality and give an example
of how the cardinality of a group of order $n$ is $\frac{1}{n}$. Some
intuition about the paths reducing the number of elements and the Euler
characteristic.

%%%%%%%%%%%%%%%%%
\subsection{Fractionals}

As an example, let's focus now on building the type
$\frac{1}{(\boolt,\ffv)}$: it will correspond to the braid group $\br_2$ on
two strands. This type will have an infinite number of points each
representing one element of the braid group $\br_2$. Mathematically, for $n$
strands, the braid group $\br_n$ is generated by $\sigma_1, \ldots,
\sigma_{n-1}$ and the following relations:
\begin{itemize}
\item $\sigma_i\sigma_{i+1}\sigma_i = \sigma_{i+1}\sigma_{i}\sigma_{i+1}$, 
\item $\sigma_i\sigma_j = \sigma_j\sigma_i$ if $|i-j| \geq 2$.
\end{itemize}
So for the particular case of $\br_2$, we have one generator $\sigma$ and no
relations. The generator corresponds to one non-trivial twist in one
direction. The group elements (i.e., the elements of
$\frac{1}{(\boolt,\ffv)}$) are the trivial braid, the $n$-fold compositions
of $\sigma$ for all $n$, and the $n$-fold compositions of $\sigma^{-1}$ for
all $n$.

So generally speaking $\eta_n$ will produce a product of $\br_n$ and $n$ and
$\epsilon_n$ will be the group action of the group on $\br_n$ on the
particular strand $n$.

%%%%%%%%%%%%%%%%%
\subsection{Negatives?}

Need operations to change the basepoint. Would these be the negatives?

@@
